\section{Steps}
\subsection{E(Expectation) Step}
\subsubsection{Introduction}In this step, we use the observed data to estimate or guess the values of the missing or incomplete data. It is used to update the variables. 
Compute the posterior probability or responsibility of each latent variable given the observed data and current parameter estimates.
Estimate the missing or incomplete data values using the current parameter estimates.
Compute the log-likelihood of the observed data based on the current parameter estimates and estimated missing data.

\subsubsection{Mathematical Formulation of E Step}
In this step we compute the expected value of  l($\theta$;X,Y).
\\\\We define,
           Q($\theta$;$\theta$_{old})=E(l($\theta$;X,Y)\(|\)X,$\theta$_{old})
\\\\ =$$\int$$l($\theta$;X,y)p(y\(|\)X,$\theta$_{old})dy


\subsection{M(Maximization) Step}
\subsubsection{Description}In this step, we need to maximize the likelihood function which we get in the E step. It can be done using several optimization numerical methods.
\subsubsection{Mathematical Formulation of M Step}

\[\theta=\underset{\theta\in\Theta}{\mathrm{arg max}}\,  Q(\theta;\theta_{old})\]
\\\\Now, l($\theta$;x)=ln(p(X\(|\)$\theta$))
\\\\=ln $\int$ p(X,y\(|\)\theta)dy
\\\\= ln $\int$ [p(X,y\(|\)\theta)p(y\(|\)X,$\theta$_{old})]/[$p(y\(|\)X,$\theta$_{old}$)] dy
\\\\=ln  E[ [(p(X,Y\(|\)$\theta$))/(P(Y\(|\)X,$\theta$_{old}))]
\(|\)X,$\theta$_{old})\hspace{1 mm}]
\\\\ \ge E \hspace{1.5 mm}[ ln \hspace{1.2 mm}[(p(X,Y\(|\)$\theta$))/p(Y\(|\)X,$\theta$_{old})]\(|\)X,$\theta$_{old})\hspace{1 mm}]
\\\\=E[ln(p(X,Y\(|\)\theta)\(|\)X,$\theta$_{old}] 
\num{-}E[ln(p(Y\(|\)X,$\theta$_{old}))\(|\)X,$\theta$_{old}]
\\\\=Q(\theta;$\theta$_{old}$)\num{-}E[ln(p(Y\(|\)X,$\theta$_{old}))\(|\)X,$\theta$_{old}]


\subsection{Derivation of the EM algorithm}

\vspace{5 mm}
\begin{itemize}
    \item Suppose we have some data X with joint pdf p(X \(|\)$\theta$)
    \item Let l($\theta$) = ln p(X\(|\)$\theta$) denotes the log - likelihood.
    \item Now suppose we are trying to guess at $\theta$ and improve our guesses through some sort of iteration. Let ${\hat{\theta}}^{(n)}$ be our nth iteration guess.
    \item We would like to find a new value for $\theta $ that satisfies: - 
    
   \hspace{35 mm}  l(${\hat{\theta}}^{(n)}$) \le l($\theta$)
    \item We will introduce some hidden variables Y, either because we are actually working with a model that has hidden(unobserved) variables or "artificially" because they make the maximization more tractable.
    \item So, we may write, 

   \hspace{20 mm} l($\theta$) - l(${\hat{\theta}}^{(n)}$) = ln p(X \(|\)$\theta$) - ln p(X \(|\)${\hat{\theta}}^{(n)}$)

   \hspace{40 mm} ln($\int$ p(X\(|\)y ,$\theta$) p(y\(|\)$\theta$) dy) - ln p(X \(|\)${\hat{\theta}}^{(n)}$)

    \hspace{41 mm}= ln($\int$ [(p(X\(|\)y,$\theta$) p(y\(|\)$\theta$))/ p(y\(|\)X, ${\hat{\theta}}^{(n)}$)]p(y\(|\)X,
    
    \hspace{43 mm} ${\hat{\theta}}^{(n)}$) dy  - ln p(X \(|\) ${\hat{\theta}}^{(n)}$)

    \item Note that the thing in parentheses is an expectation with respect to the distribution of Y \(|\)X , ${\hat{\theta}}^{(n)}$. So, applying Jensen's inequality, we have,

    \hspace{20 mm} l($\theta$) - l(${\hat{\theta}}^{(n)}$) \ge ($\int$ ln[(p(X\(|\)y,$\theta$) p(y\(|\)$\theta$))/ p(y\(|\)X, ${\hat{\theta}}^{(n)}$)]p(y\(|\)X,
    
    \hspace{43 mm} ${\hat{\theta}}^{(n)}$) dy  - ln p(X\(|\)${\hat{\theta}}^{(n)}$))

     \hspace{40 mm} = ($\int$ ln[[(p(X\(|\)y,$\theta$) p(y\(|\)$\theta$))]/ [p(y\(|\)X, ${\hat{\theta}}^{(n)}$) p(X\(|\)${\hat{\theta}}^{(n)}$)]]
     
     \hspace{43 mm} p(y\(|\)X, ${\hat{\theta}}^{(n)}$) dy)

     \ call the entire integral d($\theta$\(|\)${\hat{\theta}}^{(n)}$). 

     \ We then have , 

     \hspace{30 mm}  l(${\hat{\theta}}^{(n)}$) +  d($\theta$\(|\)${\hat{\theta}}^{(n)}$) \le l($\theta$) 
     \item Note that

     
     \hspace{20 mm} d(${\hat{\theta}}^{(n)}$\(|\)${\hat{\theta}}^{(n)}$) =$\int$ln(p(X\(|\)y,${\hat{\theta}}^{(n)})p(y\(|\)${\hat{\theta}}^{(n)}$)/(p(y\(|\)X,${\hat{\theta}}^{(n)}$)p(X\(|\)${\hat{\theta}}^{(n)}$))(p(y\(|\)X,${\hat{\theta}}^{(n)}$))dy

     \hspace{40 mm} =$\int$ln(p(X,y\(|\)$\hat{\theta}}^{(n)}$)/p(X,y\(|\)$\hat{\theta}}^{(n)}$)})p(y\(|\)X,${\hat{\theta}}^{(n)}$)dy

     \hspace{43 mm} =$\int$ln(1)p(y\(|\)X,${\hat{\theta}}^{(n)}$)dy = 0

     \item So, we have that l(${\hat{\theta}}^{(n)}$) +  d($\theta$\(|\)${\hat{\theta}}^{(n)}$) is bounded above by $l(\theta)$ and that is equal to this upper bound at the point $\theta={\hat{\theta}}^{(n)}$.
     
     If we maximize l(${\hat{\theta}}^{(n)}$) +  d($\theta$\(|\)${\hat{\theta}}^{(n)}$) with respect to \theta (equivalently \hspace{1.5 mm}maximize \hspace{1.5 mm}$d($\theta$\(|\)${\hat{\theta}}^{(n)}$)$\hspace{1.5mm} with
     
     \hspace{9 mm}respect to  $\theta$,we may improve towards maximising l($\theta$).We know at least,that
     
     \hspace{9 mm}we will not get worse.

   
      
       \hspace{4 mm} •  \hspace{0.7 mm}Maximizing 

     \hspace{20 mm} d($\theta$\(|\)${\hat{\theta}}^{(n)}$) =$\int$ln(p(X\(|\)y,$\theta$)p(y\(|\)$\theta$)/p(y\(|\)X,${\hat{\theta}}^{(n)}$)p(X\(|\)${\hat{\theta}}^{(n)}$))(p(y\(|\)X,${\hat{\theta}}^{(n)}$))dy

     \hspace{9 mm}with respect to $\theta$ is equivalent to maximizing

     \hspace{40 mm} $\int$ln(p(X\(|\)y,$\theta$)p(y\(|\)$\theta$)(p(y\(|\)X,${\hat{\theta}}^{(n)}$))dy

     \hspace{9 mm}with respect to $\theta$.
     

     \hspace{9 mm}Note that this is

     \hspace{43 mm} $\int$ln(p(X,y\(|\)$\theta$))p(y\(|\)X,${\hat{\theta}}^{(n)}$)dy
     
     \hspace{9 mm}which is 
     
     \hspace{47 mm}  E_{Y}[\ln(p(X,Y\(|\)$\theta$))\(|\)X,${\hat{\theta}}^{(n)}$]

     
 \item So, if we could compute this expectation, maximize it with respect to $\theta$ , call the result ${\hat{\theta}}^{(n+1)}$ and iterate, we can improve towards finding the $\theta$ that maximizes the likelihood (or at least not get worse).In other words , we can improve towards finding the MLE of $\theta$.

 These expectation and maximization steps are precisely the EM algorithm !

    
\end{itemize}
